MTBD models with partner notification

### 2023-07-26 article/article.tex
Partners can also notify

\usepackage{standalone}
\usepackage{caption}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\usepackage{tikzsymbols}
\usepackage{hieroglf}
\usepackage{nameref}
\usepackage{algpseudocode}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}

@@

Another important source of heterogeneity that none of these models account for is inflicted by non-random sampling. We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for non-randomness of sampling, in particular due to contact tracing and partner notification.

In this study we propose an extension of the MTBD framework that allows for modeling non-random sampling due to partner notification (PN), MTBD-PN. In the rest of the article, we describe the MTBD-PN model family and its assumptions, present a transmission tree simulator under these models, derive the equations for MTBD-PN likelihood calculation, and propose a test for detecting partner notification in pathogen phylogenetic trees. We also present a maximum likelihood parameter estimator for the simplest model of the MTBD-PN family, BD-PN, and test it on simulated data. Finally, we apply it to the study of HIV-1 B epidemic in the UK.

\section{MTBD-PN Framework}
In a pathogen transmission tree $\mathscr{T}$ (approximated by a time-scaled pathogen phylogeny) the tips represent sampled pathogens, patient state transitions occur along the branches, and bifurcations (i.e. internal nodes) correspond to pathogen transmissions (Fig.~\ref{fig:tt}). The tree branches are measured in units of time, where $T$ is the time that passed between the tree root (the beginning of the (sub-)epidemic) and the last sampled tip.

\begin{figure*}[tbhp]
\centering 
%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
\includestandalone[width=1\textwidth]{Fig_tree}
\caption{A transmission tree $\mathscr{T}$ with $n=5$ external nodes (i.e. tips, which correspond to sampling events: $0000, 0001, 001, 010, 011$), $n-1=4$ internal nodes (which correspond to transmissions: $0$ (the root) and $00, 01, 000$) and $2n - 2 = 8$ branches (plus the root branch of zero length). %The lengths of the branches are shown on the right of each branch: $t_i$ is the lengths of the branch connecting the node $i$ to its parent. 
Time $t$ starts at the root of the tree ($t=0$) and goes till the last sampled tip. The times of the nodes are shown on the left, e.g. $t_{0001}$ is the time of tip $0001$ (when $0001$'s pathogen was sampled). $T$ corresponds to the end of the sampling period (when the most recent tip, $0000$, was sampled).}
\label{fig:tt} 
\end{figure*}

\subsection{MTBD Framework}
A general MTBD model describes $m$ possible individual states. An individual in state $k \in 1:m$ can be removed (i.e., become non-infectious, e.g., due to healing, starting a treatment, self-isolation, or death) at a rate $\psi_k$, change their state to a state $l \in 1:m$ at a rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in a state $l \in 1:m$ at rate $\lambda_{kl}$. Upon $i$'s removal, their pathogen can be sampled (and hence observed as a tip in the transmission tree) with a probability $\rho$.

The MTBD model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$). These likelihood densities depend on the probability $U_s(t)$ of an unobserved transmission tree that started from one individual at time $t$ in state $s \in 1:m$ and evolved till $T$, without them or any of their induced cases being sampled:

\begin{equation}
\scriptsize
\begin{cases}
\dot{U}_{s}(t) = &\Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) U_{s}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\ 
    & - \sum\limits_{k=1}^{m} \mu_{sk} U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by unsampled evolution from the new state }\\
    &- \sum\limits_{k=1}^{m} \lambda_{sk} U_{s}(t)U_{k}(t) \;  \textit{\color{gray} $\leftarrow$ transmission, followed by unsampled evolutions of both subtrees}\\
    &- \psi_s (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
U_{s}(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:Us}
\end{cases}
\end{equation}

With the equations~(\ref{eq:Us}), we can write down the equations for the probability densities $p_{s,si}^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$ in state $si$, starting  on this branch at a time $t \leq t_i$ in state $s$. The master equations for the MTBD models were initially developed by ~\citet{Stadler2013a}, however what we present here, is their parallelizable version that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_{s,si}^{(i)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, folowed by evolution from the new state }\\
    & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{s,si}^{(i)}(t)U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
    & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{k,si}^{(i)}(t)U_{s}(t)\;  \textbf{\textit{\color{gray} $\leftarrow$ transmission, where the donor subtree stayed unsampled}}\\
p_{s,si}^{(i)}(t_i) = & 
    \begin{cases}
    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
    1 & \textit{\color{gray} if $i$ is in state $si$}
    \end{cases}\label{eq:p}
\end{cases}
\end{equation}

If the states of all the observed nodes of the tree $\mathscr{T}$ are known, then its likelihood density under parameter values $\Theta$ can be calculated as $L(\mathscr{T}|\Theta)$:

{\scriptsize
\begin{alignat*}{3}
log L(\mathscr{T}|\Theta) =  \sum\limits_{i \in tips}  log(&\psi_{si}  \rho) & \textit{\color{gray} $\leftarrow$ sampling of $n$ tips} \\
 +\sum\limits_{i \in \textit{internal nodes}} log \Big(&\sum\limits_{k=1}^{m} \lambda_{si,k}  & \textit{\color{gray} $\leftarrow$ $n - 1$ transmission events}\\
 & \cdot \big(p_{si,si0}^{(i0)}(t_i)p_{k,si1}^{(i1)}(t_i) + p_{k,si0}^{(i0)}(t_i)p_{si,si1}^{(i1)}(t_i)\big)\Big) & \textit{\color{gray}$\leftarrow$ child branch evolutions}  \label{eq:mtbd_tree_likelihood}
\end{alignat*}}

@@

We now propose a non-Markovian extension of the MTBD framework that adds partner notification (PN), for a tree with known node states.  In MTBD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a given probability $\rho_n$. Upon notification, partner is removed almost instantaneously (modeled via a notified removal rate $\psi_n >> \psi_k \;\forall k \in 1:m$). MTBD-PN hence has $2k^2 + 1$ classical MTBD and 2 additional parameters:
\begin{itemize}
 \item $\mu_{kl} \;\forall 1 \leq k,l  \leq m, k \neq l$ -- constant state change rate from $k$ to $l$;
 \item $\lambda_{kl} \;\forall 1 \leq k,l \leq m$ -- constant transmission rate: an infected individual in state $k$ transmits their pathogen (which corresponds to an internal node in the transmission tree), the newly infected individual is in state $l$;
 \item $\psi_k \;\forall 1 \leq k \leq m$ -- constant removal rate: an infected individual in state $k$ stops being infectious at this rate. This corresponds to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
 \item $\rho_n$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
 \item $\psi_n >> \psi_k  \;\forall 1 \leq k \leq m$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate. This creates an observed tip in the transmission tree.
\end{itemize}

This model is a simplification of the notification process, in particular it makes 5 assumptions:
\begin{enumerate}
\item only observed individuals can notify (instead of any removed individual);
\item notified individuals are always observed upon removal;
\item notified individuals do not notify further;
\item after the notification, notified individuals do not transmit further;
\item only the most recent partner can get notified.
\end{enumerate}

@@

\subsubsection{Mathematical formulation}
$p_{s,si}^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along a tree branch, provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$. These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.

We will use an \textbf{oriented probability} $p^{(i,o)}_{s,si}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_{s,si}^{(i,o)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,o)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, followed by evolution from the new state }\\
    & - \sum\limits_{k=1}^{m} \lambda_{sk} p_{s,si}^{(i,o)}(t)U_{k}(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
p_{s,si}^{(i,o)}(t_i) = & 
    \begin{cases}
    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
    1 & \textit{\color{gray} if $i$ is in state $si$}
    \end{cases}\label{eq:p-o}
\end{cases}
\end{equation}

Additionally we define a \textbf{non-hidden probability} $p^{(i,nh)}_{s,si}(t)$ for the cases when no hidden transmission occurred along the branch:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}_{s,si}^{(i,nh)}(t) = & \Big(\sum\limits_{k=1}^{m} \mu_{sk} + \sum\limits_{k=1}^{m} \lambda_{sk} + \psi_s\Big) p_{s,si}^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \sum\limits_{k=1}^{m} \mu_{sk} p_{k,si}^{(i,nh)}(t)\;  \textit{\color{gray} $\leftarrow$ change of the state, followed by evolution from the new state }\\
p_{s,si}^{(i,nh)}(t_i) = & 
    \begin{cases}
    0 & \textit{\color{gray} if $i$ is in state $k \neq si$}\\
    1 & \textit{\color{gray} if $i$ is in state $si$}
    \end{cases}\label{eq:p-nh}
\end{cases}
\end{equation}

Under the MTBD-PN model, the probability density $q_{si,sj}(i)$ of a branch connecting a node $i$ in state $si$ to its parent node $j$ in state $sj$ (without taking into account the event at the end of the branch) can be written as:

\begin{equation}
\scriptsize
q_{sj,si}(i) = 
\begin{cases}
q^{(u)}_{sj,si}(i) = p^{(i)}_{sj,si}(t_j) \textit{\color{gray} ~if $i$ is an \textbf{unnotified non-notifier}}\\
q_{sj,si}^{(n)}(i) = 
p_{sj,si}^{(i,nh)}(t_j) \textit{\color{gray} ~~if $i$ is a tip corresponding to a \textbf{notifier}}\\
~~~~~~~~~~~~ \textit{\color{gray}(parent is the last transmission, no hidden transmission along the branch)}
\\
q^{(p)}_{sj,si}(i)  = p^{(i, o)}_{sj,si}(t_j)  \textit{\color{gray} ~if $i$ is an internal node whose individual will be notified after $t_i$}
\\
q^{(p)}_{sj,si}(i, r)  = 
\begin{cases}
p^{(r, o)}_{sj,si}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r \leq t_i$}\\
p^{(i, o)}_{sj,si}(t_j) \textit{\color{gray}~~ if $i$ is a \textbf{partner} tip sampled via a standard procedure} \\
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\textit{\color{gray} before the notification by $r$: $t_i < t_r$}\\
\end{cases}
\end{cases}
\label{eq:branch}
\end{equation}

Using the formulas~(\ref{eq:branch}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}, where for each visited node $i$, and for each possible state $1 \leq sj \leq m$ at the beginning of the branch connecting it to its parent $j$  we calculate the likelihood density of its subtree if its branch corresponds to (1) an unnotified individual $l^{(u)}_{sj}(i)$; (2) a notifier $l^{(n)}_{sj}(i)$ (for tips only); or (3) an (eventually) notified partner $l^{(p)}_{sj}(i, r)$ (notified by the individual corresponding to the tip $r$ at time $t_r$).

For a tip $i$ in state $si$ these values can be calculated as:
\begin{equation}
\scriptsize
\begin{array}{ll}
l^{(u)}_{sj}(i) &= q^{(u)}_{sj}(i)\psi_{si}\rho (1-\rho_{n})\\
l^{(n)}_{sj}(i) &= q^{(n)}_{sj}(i)\psi_{si}\rho \rho_{n}\\
l^{(p)}_{sj}(i, r) &= q^{(p)}_{sj}(i,r)\psi_{n}\\
\end{array}
\label{eq:tip}
\end{equation}

Let us now consider an internal node $i$ in state $si$ with two child nodes, $i0$ and $i1$. We will start with the configuration where the $i$'s branch corresponds to an unnotified individual:

\begin{equation}
\scriptsize
\begin{split}
l^{(u)}_{sj}(i) = &q^{(u)}_{sj}(i) \sum\limits_{k=1}^{m} \lambda_{si,k} \\
\cdot \Big(~&
l^{(u)}_{si}(i0)l^{(u)}_{k}(i1) + l^{(u)}_{k}(i0)l^{(u)}_{si}(i1)  \textit{\color{gray} $\leftarrow$ no notification on child branches} \\
& + is\_tip(i0) \big(l^{(n)}_{si}(i0)l^{(p)}_{k}(i1,i0) + l^{(n)}_{k}(i0)l^{(p)}_{si}(i1,i0) \big)\textit{\color{gray} $\leftarrow$ notification of someone in $i1$'s subtree by $i0$} \\
&+ is\_tip(i1) \big(l^{(p)}_{si}(i0,i1)l^{(n)}_{k}(i1) + l^{(p)}_{k}(i0,i1)l^{(n)}_{si}(i1) \big)~\Big) \textit{\color{gray} $\leftarrow$ notification of someone in $i0$'s subtree by $i1$},\\
\textit{where } & is\_tip(r) = 1 \textit{ if r is a tip and $= 0$ if r is an internal node.}
\end{split}\label{eq:lu}
\end{equation}

Note that a notification will only be possible if the notifying branch is external as the notifier only notifies the most recent partner, i.e. the transmission corresponds to the notifier's parent node. %; (2) there is at least one tip in the notified subtree that can be a partner, i.e. was sampled after the notifier and whose parent's time was before notification (i.e. there was no transmission between notification and sampling).

In the other case (when an internal nodes's branch corresponds to a partner who will eventually get notified by a node $r$), the individual on the path connecting this branch to the notified tip stays the same (the partner). Hence, all the transmissions on this path are oriented and correspond to the partner transmitting to another recipient individual. Moreover, this recipient's branch cannot be notified by our partner (as in our model notified partners do not notify further):

\begin{equation}
\scriptsize
\begin{split}
l^{(p)}_{sj}(i, r) = q^{(p)}_{sj}(i,r) \sum\limits_{k=1}^{m}  \lambda_{si,k}\Big(&l^{(p)}_{si}(i0,r)l^{(u)}_{k}(i1) \textit{\color{gray} $\leftarrow$ notification in $i0$'s subtree} \\
 &+ l^{(u)}_{k}(i0)l^{(p)}_{si}(i1,r)\Big)  \textit{\color{gray} $\leftarrow$ notification in $i1$'s subtree} 
\end{split}\label{eq:lu}
\end{equation}

Note that $l^{(p)}_{sj}(i, r)$ depends on the notifier $r$ and can only be calculated once the notification time is known, i.e. once we have climbed to an internal node corresponding to the transmission between the notifier (corresponding to one of its children, must be a tip) and the partner (the other child). In the partner subtree only the contribution coming from the partner tip branch depends on the notification time. Hence while climbing the tree from tips to the root, for each internal node $i$ and each of the possible states $sj \in 1:m$ at the beginning of its branch, we can store an array of pre-calculated values for notified internal paths leading to each of its tips. The values of this array will store for each tip (e.g. $a$) its sampling time $t_a$ and a configuration $C_{sj}(i,a)$, representing the subtree likelihood provided $a$ corresponds to the same notified partner individual as $i$ (but without the $a$'s tip branch contribution):

>>>>

\usepackage{standalone}
\usepackage{caption}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\usepackage{tikzsymbols}
\usepackage{hieroglf}
\usepackage{nameref}
\usepackage{algpseudocode}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{stackengine}

@@

Another important source of heterogeneity that none of these models accounts for is inflicted by non-random sampling. We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for non-randomness of sampling, in particular due to contact tracing and partner notification.

In this study we propose an extension of the BD model that allows for modeling non-random sampling due to partner notification (PN), BD-PN. In the rest of the article, we describe the BD-PN model and its assumptions, present a transmission tree simulator under this model, derive the equations for BD-PN likelihood calculation, and propose a test for detecting partner notification in pathogen phylogenetic trees. We also present a maximum likelihood parameter estimator for BD-PN, and test it on simulated data. Finally, we apply it to the study of HIV-1 B epidemic in the UK.

\section{BD-PN Framework}
In a pathogen transmission tree $\mathscr{T}$ (approximated by a time-scaled pathogen phylogeny) the tips represent sampled pathogens %, patient state transitions occur along the branches, 
while bifurcations (i.e. internal nodes) correspond to pathogen transmissions (Fig.~\ref{fig:tt}). The tree branches are measured in units of time, where $T$ is the time that passed between the tree root (the beginning of the (sub-)epidemic) and the last sampled tip.

\begin{figure*}[tbhp]
\centering 
%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
\includestandalone[width=0.6\textwidth]{Fig_tree}
\caption{A transmission tree $\mathscr{T}$ with $n=5$ external nodes (i.e. tips, which correspond to sampling events: $0000, 0001, 001, 010, 011$), $n-1=4$ internal nodes (which correspond to transmissions: $0$ (the root) and $00, 01, 000$) and $2n - 2 = 8$ branches (plus the root branch of zero length). %The lengths of the branches are shown on the right of each branch: $t_i$ is the lengths of the branch connecting the node $i$ to its parent. 
Time $t$ starts at the root of the tree ($t=0$) and goes till the last sampled tip. The times of the nodes are shown on the left, e.g. $t_{0001}$ is the time of tip $0001$ (when $0001$'s pathogen was sampled). $T$ corresponds to the end of the sampling period (when the most recent tip, $0000$, was sampled).}
\label{fig:tt} 
\end{figure*}

\subsection{BD Model with incomplete sampling}
Under the basic birth-death model~\citep{Stadler2009} all the individuals are in the same state. They can transmit with a constant rate $\lambda$, get removed with a constant rate $\psi$, and their pathogen can be sampled upon removal with a constant probability $\rho$.

%A general MTBD model describes $m$ possible individual states. An individual in state $k \in 1:m$ can be removed (i.e., become non-infectious, e.g., due to healing, starting a treatment, self-isolation, or death) at a rate $\psi_k$, change their state to a state $l \in 1:m$ at a rate $\mu_{kl}$ (where $\mu_{kk} = 0$), and transmit their pathogen to an individual in a state $l \in 1:m$ at rate $\lambda_{kl}$. Upon $i$'s removal, their pathogen can be sampled (and hence observed as a tip in the transmission tree) with a probability $\rho$.

This model can be described with master equations representing the likelihood densities of evolving as on the observed transmission tree, with time $t$ going forward from the root ($t=0$) till the time of the last sampled tip ($t=T$). These likelihood densities depend on the probability $U(t)$ of an unobserved transmission tree that started from one individual at time $t$ and evolved till $T$, without them or any of their induced cases being sampled:

\begin{equation}
\scriptsize
\begin{cases}
\dot{U}(t) = &\Big(\lambda + \psi\Big) U(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    &- \lambda U^2(t) \;  \textit{\color{gray} $\leftarrow$ transmission, followed by unsampled evolutions of both subtrees}\\
    &- \psi (1 - \rho)\;  \textit{\color{gray} $\leftarrow$ removal without sampling}\\
U(T) = & 1\;  \textit{\color{gray} $\leftarrow$ the probability to stay unsampled over time 0 is 1} \label{eq:Us}
\end{cases}
\end{equation}

With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch that finishes at time $t_i$, starting on this branch at a time $t \leq t_i$. The master equations for the BD models were initially developed by ~\citet{Stadler2009}, however what we present here, is their branch-specific formulation that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}^{(i)}(t) = & \Big(\lambda + \psi\Big) p^{(i)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - 2 \lambda p^{(i)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where one of the subtrees stayed unsampled}\\
p^{(i)}(t_i) = 1
\end{cases}\label{eq:p}
\end{equation}

Using this equation, we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ as:

\begin{equation}
\scriptsize
\begin{split}
log L(\mathscr{T}|\Theta) =  &\sum\limits_{i \in tips}  log(\psi \rho)  \textit{\color{gray} $\leftarrow$ sampling of $n$ tips} \\
 &+\sum\limits_{i \in \stackanchor{\tiny\textit{internal}}{\tiny\textit{nodes}}} log \big(2 \lambda p^{(i0)}(t_i)p^{(i1)}(t_i)\big)   \textit{\color{gray} $\leftarrow$ $n - 1$ transmission events,}\\
 & ~~~~~~~~~~~~~~~\textit{\color{gray} followed by child branch evolutions (each can be a donor)}  \label{eq:mtbd_tree_likelihood}
\end{split}
\end{equation}

The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g. by finding parameter values that maximize it, in maximum likelihood framework). In particular, it permits inference of the following epidemiological parameters:

\begin{itemize}
\item \textit{effective reproduction number} $R_e = \frac{\lambda}{\psi}$, expected number of individuals directly infected by an infectious case;
\item \textit{infectious time} $\frac{1}{\psi}$, time during which an infectious individual can further spread the epidemic.
\end{itemize}

Importantly, the BD model is asymptotically unidentifiable (see Remark 3.4 in~\citep{Stadler2009}), but to become identifiable it requires one of the parameters to be fixed. In practice, it is often the sampling probability, as it may be approximated from epidemiological data $\rho$ (e.g., the proportion of sampled cases among the declared ones) or the infectious time $\frac{1}{\psi}$ (estimated from observations of infected cases).

@@

We now propose a non-Markovian extension of the BD model that adds partner notification (PN).  In BD-PN, at the moment of sampling the sampled individual might notify their most recent partner with a given probability $\rho_n$. Upon notification, partner is removed almost instantaneously (modeled via a notified removal rate $\psi_n >> \psi$). BD-PN hence has 3 classical BD and 2 additional parameters:
\begin{itemize}
 \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual (which corresponds to an internal node in the transmission tree);
 \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This corresponds to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
 \item $\rho_n$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
 \item $\psi_n >> \psi$ -- constant notified removal rate: a notified individual (if not yet removed via standard procedure) stops being infectious and gets observed with this rate. This creates an observed tip in the transmission tree.
\end{itemize}

This model is a simplification of the notification process, in particular it makes 4 assumptions:
\begin{enumerate}
\item only observed individuals can notify (instead of any removed individual);
\item notified individuals are always observed upon removal;
%\item notified individuals do not notify further;
\item after the notification, notified individuals do not transmit further;
\item only the most recent partner can get notified.
\end{enumerate}

@@

Note that the most recent partner relationship is not necessarily symmetrical, as the most recent partner of $i$ might have transmitted their virus further after the contact with $i$.

\subsubsection{Mathematical formulation}
$p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along a tree branch, provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$. These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.

We will use an \textbf{oriented probability} $p^{(i,o)}(t)$ for the cases when the individual at the end of the branch is the same as at time $t$:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}^{(i,o)}(t) = & \Big(\lambda + \psi\Big) p^{(i,o)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
    & - \lambda p^{(i,o)}(t)U(t)\;  \textit{\color{gray} $\leftarrow$ transmission, where the recipient subtree stayed unsampled}\\
p^{(i,o)}(t_i) = 1\label{eq:p-o}
\end{cases}
\end{equation}

Additionally we define a \textbf{non-hidden probability} $p^{(i,nh)}(t)$ for the cases when no hidden transmission occurred along the branch:

\begin{equation}
\scriptsize
\begin{cases}
\dot{p}^{(i,nh)}(t) = & \Big(\lambda + \psi\Big) p^{(i,nh)}(t)\; \textit{\color{gray} $\leftarrow$ no event in the next infinitesimal time $\Delta t$ }\\
p^{(i,nh)}(t_i) = 1\label{eq:p-nh}
\end{cases}
\end{equation}

The equations~\ref{eq:Us}, \ref{eq:p}, \ref{eq:p-o}, and \ref{eq:p-nh} have the following closed form solutions:
\begin{equation}
\scriptsize
\begin{split}
&\begin{cases}
U(t) = \frac{\lambda + \psi}{2\lambda} +  \frac{c_1}{2\lambda}\Big(\frac{E_1 - 1}{E_1 + 1}\Big)\\
\\
p^{(i)}(t) = \Big(\frac{E_2 + 1}{E_1 + 1}\Big)^2E_3 \\
\\
p^{(i,o)}(t) =  \frac{E_2 + 1}{E_1 + 1}\sqrt{E_3E_4} \\
\\
p^{(i,nh)}(t) =  E_4
\end{cases},\\
& \textit{where } c_1 = \sqrt{(\lambda - \psi)^2 + 4 \lambda\psi\rho},\; c_2 = \frac{c_1 + \lambda - \psi}{c_1 - \lambda + \psi},\\
& ~~~~~~~~E_1 = c_2 e^{c_1 (t -T)},\; E_2 = c_2 e^{c_1 (t_i -T)},\; E_3 = e^{c_1 (t -t_i)}, \; E_4 =  e^{(\lambda + \psi) (t -t_i)}\\
\end{split}\label{eq:ps}
\end{equation}

%Under the BD-PN model, the probability density $q(i)$ of a branch connecting a node $i$ to its parent node $j$ (without taking into account the event at the end of the branch) can be written as:
%
%
%\begin{equation}
%\scriptsize
%q(i) = 
%\begin{cases}
%q^{(u)}(i) = p^{(i)}(t_j) \textit{\color{gray} ~if $i$ is an \textbf{unnotified non-notifier}}\\
%q^{(n)}(i) = 
%p^{(i,nh)}(t_j) \textit{\color{gray} ~~if $i$ is a tip corresponding to a \textbf{notifier}}\\
%~~~~~~~~~~~~ \textit{\color{gray}(parent is the last transmission, no hidden transmission along the branch)}
%\\
%q^{(p)}(i)  = p^{(i, o)}(t_j)  \textit{\color{gray} ~if $i$ is an internal or tip node whose individual will be notified after $t_i$}
%\\
%q^{(p)}(i, r)  = p^{(r, o)}(t_j) e^{- \psi_n (t_i - t_r)} \textit{\color{gray} if $i$ is a \textbf{partner} tip notified by $r$ at time $t_r \leq t_i$}
%\end{cases}
%\label{eq:branch}
%\end{equation}

Using the formulas~(\ref{eq:ps}), we can calculate the likelihood density of a tree using a pruning algorithm~\cite{10.1093/sysbio/22.3.240}, where for each visited node $i$ we calculate the likelihood density of its subtree if the branch connecting it to its parent node $j$ corresponds to (1) an unnotified individual $l^{(u)}(i)$; or (2) an (eventually) notified partner $l^{(p)}(i, r)$ (notified by the individual corresponding to the tip $r$ at time $t_r$). For each visited tip we additionally consider whether the corresponding individual notified their most recent partner or not.

%For a tip $i$ in state $si$ these values can be calculated as:
%\begin{equation}
%\scriptsize
%\begin{array}{ll}
%l^{(u)}(i) &= \begin{cases} l^{(u,n)}(i) = q^{(u)}(i)\psi\rho
%\rho_{n} \textit{\color{gray} ~~~~~~~~if $i$ notified their most recent patner}\\
%l^{(u,\not{n})}(i) = q^{(u)}(i)\psi\rho (1-\rho_{n}) \textit{\color{gray} ~if $i$ did not notify}
%\end{cases}\\
%l^{(p)}(i, r) &= q^{(p)}(i,r)\psi_{n}\cdot \begin{cases}
%\rho_{n} \textit{\color{gray} ~~~~~~~~if $i$ notified their most recent patner}\\
%(1-\rho_{n}) \textit{\color{gray} ~if $i$ did not notify}
%\end{cases}\\
%\end{array}
%\label{eq:tip}
%\end{equation}

For a tip $i$ with a parent node $j$ these values can be calculated as:
\begin{equation}
\scriptsize
\begin{array}{ll}
l^{(u,n)}(i) &= p^{(i,nh)}(t_j)\psi\rho
\rho_{n} \textit{\color{gray} ~if $i$ was not notified but notified their most recent patner}\\
l^{(u,\not{n})}(i) &= p^{i}(t_j)\psi\rho (1-\rho_{n}) \textit{\color{gray} ~if $i$ was not notified and did not notify}\\
l^{(p,n)}(i,r) &= \begin{cases}
p^{(i, nh)}(t_j)\psi\rho\rho_{n} \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and notified their most recent patner}\\
p^{(r, nh)}(t_j) e^{- \psi_n (t_i - t_r)} \psi_n \rho_{n} \textit{\color{gray} ~if $i$ got notified at time $t_r \leq t_i$ and notified}
\end{cases}\\
l^{(p,\not{n})}(i,r) &= \begin{cases}
p^{(i, o)}(t_j)\psi\rho(1 - \rho_{n}) \textit{\color{gray} ~if $i$ got notified at time $t_r > t_i$ and did not notify}\\
p^{(r, o)}(t_j) e^{- \psi_n (t_i - t_r)} \psi_n (1-\rho_{n}) \textit{\color{gray} ~if $i$ got notified at time $t_r \leq t_i$ and did not notify}
\end{cases}\\
\end{array}
\label{eq:tip}
\end{equation}

Let us now consider an internal node $i$ with two child nodes, $i_0$ and $i_1$. We will start with the configuration where the $i$'s branch corresponds to an unnotified individual. We will consider three possibilities: (i) when both $i_0$ and $i_1$ are internal nodes; (ii) when both $i_0$ and $i_1$ are tips; and (iii) when one of them is an internal node and the other one is a tip.

%Note that a notification will only be possible if the notifying branch is external as the notifier only notifies the most recent partner, i.e. the transmission corresponds to the notifier's parent node.

First, let's assume that (i) $i_0$ and $i_1$ are internal. Then none of them can be a notifier (since notifiers correspond to tips), and hence none of them is notified:

\begin{equation}
\scriptsize
l^{(u)}(i) = p^{i}(t_j) 2\lambda l^{(u)}(i_0)l^{(u)}(i_1) \label{eq:lu-int-int}
\end{equation}

Secondly, let's assume that (ii) $i_0$ and $i_1$ are tips. Then each of them could have notified the other one:

\begin{equation}
\scriptsize
\begin{split}
l^{(u)}(i) = p^{i}(t_j) 2\lambda \cdot 
\Big(&l^{(u,\not{n})}(i_0)l^{(u,\not{n})}(i_1) \textit{\color{gray}~$\leftarrow$ neigher $i_0$ nor $i_1$ notified} \\
&+ l^{(u,n)}(i_0)l^{(p,\not{n})}(i_1,i_0) \textit{\color{gray}~$\leftarrow$ $i_0$ notified $i_1$ and $i_1$ did not notify} \\
&+ l^{(u,n)}(i_1)l^{(p,\not{n})}(i_0,i_1) \textit{\color{gray}~$\leftarrow$ $i_1$ notified $i_0$ and $i_0$ did not notify} \\
&+ l^{(p,n)}(i_0,i1)l^{(p,n)}(i_1,i_0)\Big) \textit{\color{gray}~$\leftarrow$ $i_1$ and $i_0$ notified each other} \\
\end{split}\label{eq:lu-tip-tip}
\end{equation}
